\chapter*{Abstract}

This paper presents IncidentFlow, a domain-specific language (DSL) for cybersecurity incident response documentation and automation. In prevailing practice, response procedures are maintained as static documents and executed manually under severe time constraints, forcing teams to repeatedly check mandatory sections, formatting constraints, and compliance evidence while incidents are still unfolding. As established in Chapter 1, this operational pattern contributes to procedural omissions, coordination latency, and weakened auditability.

The core problem is that incident response playbooks remain largely static, difficult to validate before real incidents, and disconnected from executable workflows. Existing automation ecosystems provide partial remediation but are frequently constrained by vendor lock-in, elevated adoption cost, and limited support for transparent version-controlled review workflows. Consequently, organizations lack a practical mechanism that is simultaneously human-readable for responders and machine-executable for dependable automation.

The proposed solution encodes incident logic and reporting semantics in a unified formal specification, including triggers, phases, control-flow constructs, verification handlers, severity transitions, and reporting actions. This design reduces repetitive requirement-checking by making structural completeness and procedural consistency properties of the language model itself.

Implementation is realized through an ANTLR-based parsing pipeline comprising a rigorously specified grammar, domain-constrained lexical analysis, and deterministic parser behavior with explicit error instrumentation, as detailed in Chapter 2. This architecture furnishes a syntactically validated intermediate representation that can be leveraged for downstream automation, conformance assurance, and reproducible report synthesis.

The principal contribution is a methodologically robust and extensible foundation for process-aware incident reporting that improves operational efficiency, attenuates human error, and strengthens evidentiary quality in high-pressure security operations.

\textbf{Keywords: } Domain-Specific Languages, Cybersecurity Incident Response, Incident Reporting Automation, ANTLR, Parser Design, Compliance Engineering, Human Error Reduction.
